``Interestingness'' in Mathematical Theory Exploration

### 2019-10-04 PhDThesis/related_formalstats.tex
Fix listings
Incorporate feedback

\subsection{Statistics of Formal Systems}

The core problem of assigning ``interestingness'' to logical formulae is the
application of statistical reasoning to the discrete, semantically-rich domain
of formal systems. This problem has been tackled from various directions for a
variety of reasons; here we summarise those contributions which seem of
particular importance for theory exploration.

\subsubsection{Clustering}
\label{sec:clustering}

% TODO: ML4PG
\cite{journals/corr/abs-1212-3618}
% TODO: ACL2(ml) (also examples section)
\cite{heras2013proof}

\subsubsection{Probability of Sentences}

The most important property of a logical formula is its truth value. Although we
may be able to determine some truth values exactly, e.g. using decision or
semi-decision procedures, it may be more efficient to \emph{approximate} truth
values. One straightforward extension of truth values is \emph{probabilities},
where we can assign probability $1$ to formulae which are known to be true, $0$
to formulae known to be false, and intermediate values to those which we do not
yet know.

% TODO
\cite{Hutter.Lloyd.Ng.ea:2013}

\subsubsection{Interestingness in Concept Formation}
\label{sec:conceptformation}

@@

% Eurisko, AM, etc.?

\subsubsection{Learning From Structured Data}

@@

\paragraph{Feature Extraction}\label{sec:featureextraction}

@@

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
data Maybe a = Nothing | Just a

data Either a b = Left a | Right b
\end{lstlisting}

@@

\paragraph{Truncation and Padding}

@@

\paragraph{Dimension Reduction}

A more sophisticated approach to the problem of reducing input size is to view
it as a \emph{dimension reduction} technique: our inputs can be modelled as
points in high-dimensional spaces, which we want to project into a
lower-dimensional space ($\left\{ {0, 1} \right\}^N$ in the case of $N$-bit
vectors).

Truncation is a trivial dimension reduction technique: take the first $N$
coordinates (bits). More sophisticated projection functions consider the
\emph{distribution} of the points, and project with the hyperplane which
preserves as much of the variance as possible (or, equivalently, reduces the
\emph{mutual information} between the points).

There are many techniques to find these hyperplanes, such as \emph{principle
  component analysis} (PCA) and \emph{autoencoding}; however, since these
techniques are effectively ML algorithms in their own right, they suffer some of
the same constraints we're trying to avoid:

@@

\paragraph{Sequences}

To handle input of \emph{variable} size, most attention has been given to the
handling of \emph{sequences}. This is a lossless approach, which splits the
input into fixed-size \emph{chunks}, which are fed into an appropriate ML
algorithm one at a time. The sequence is terminated by a sentinel; an
``end-of-sequence'' marker which, by construction, is distinguishable from the
data chunks. This technique allows us to trade \emph{space} (the size of our
input) for \emph{time} (the number of chunks in a sequence).

@@

\paragraph{Recursive Structure}

@@

The application of \emph{kernel methods} to structured information is discussed
in \cite{Gartner2003}, where the input data (including sequences, trees and
graphs) are represented using \emph{generative models}, such as hidden Markov
models, of a fixed size.

@@

% TODO
\cite{Gartner2003}
\cite{Oveisi.Oveisi.Erfanian.ea:2012}
\cite{bakir2007predicting}
\cite{conf/ijcai/Plate91}
\cite{goller1996learning}
\cite{kwasny1995tail}
\cite{pollack1990recursive}
\cite{zanzotto2012distributed}

>>>>

\section{Statistics of Formal Systems}

\iffalse
TODO: Alison: I'd like to revisit this chapter once I see the whole
thesis. At the moment it feels in parts more like a discussion of your work,
than of related work. It might help to read other people's related work sections
to get a better idea of what to include here. I think it'll be more obvious too,
once we have the intro, etc. I suspect you'll need to do a fair bit more work on
this one.
\fi

The core problem of assigning ``interestingness'' to logical formulae is the
application of statistical reasoning to the discrete, semantically-rich domain
of formal systems. This problem has been tackled from various directions for a
variety of reasons; here we summarise those contributions which are of
particular importance for theory exploration.

\iffalse
\subsection{Clustering}
\label{sec:clustering}

% TODO: ML4PG
\cite{journals/corr/abs-1212-3618}
% TODO: ACL2(ml) (also examples section)
\cite{heras2013proof}
\fi

\subsection{Probability of Statements}

An important property of a logical formula (in general, and in particular with
regards to interestingness) is its truth value. Although we may be able to
determine some truth values exactly, e.g. via deduction, it may only be possible
(or more efficient) to \emph{approximate} truth values. One straightforward
approximation is to adopt \emph{probabilities}, where we assign probability $1$
to formulae which are known to be true (e.g. axioms), $0$ to formulae known to
be false, and intermediate values to those which are not known.

This can be useful for ``knowledge bases'' of non-logical axioms (e.g. the
observed colours of Hempel's ravens~\cite{10.1093/mind/LIV.214.97} \iffalse
TODO: Alison: Maybe a bit of explanation here \fi), and allows the probability of
universal statements to be inferred from non-exhaustive
cases~\cite{Hutter.Lloyd.Ng.ea:2013}. However, inferring consistent
probabilities requires \emph{logical omniscience}: knowledge of all the logical
consequences of the chosen axioms. This is unfortunate, since it is precisely
what we are trying to avoid by approximating with probabilities.

Avoiding this requirement for logical omniscience has given rise to the field of
\emph{bounded rationality}, where reasoning (deduction, induction, etc.) is
subject to some computational constraints. For example, the ``logical inductor''
of~\cite{garrabrant2016logical} avoids assigning particular probabilities to
formal statements, and instead produces a \emph{sequence} of successive
approximate probabilities, each computable with finite (though
doubly-exponential!) resources. These approximations are only guaranteed to be
consistent in the limit.

The probability of a statement can affect its interestingness in a few ways.
Statements whose probability is very close to $0$ or $1$ may be considered
less interesting, since they are essentially known (e.g. easily derivable or
refutable via known facts). Those closer to $\frac{1}{2}$ may be considered more
interesting; although in a bounded rationality context this may simply indicate
the requirement for some routine (but lengthy) computation: for example stating
that the $n$th bit of $\pi$ is a zero, for some large $n$.

\iffalse
\subsection{Interestingness in Concept Formation}
\label{sec:conceptformation}

@@

% Eurisko, AM, etc.?
\fi

\subsection{Learning From Structured Data}

@@

\subsubsection{Feature Extraction}\label{sec:featureextraction}

@@

\begin{haskell}
data Maybe a = Nothing | Just a

data Either a b = Left a | Right b
\end{haskell}

@@

\subsubsection{Truncation and Padding}

@@

\subsubsection{Dimension Reduction}

A more sophisticated approach to the problem of reducing input size is to view
it as a \emph{dimension reduction} technique: our inputs can be modelled as
points in high-dimensional spaces, which we want to project into a
lower-dimensional space.

Truncation is a trivial dimension reduction technique: take the first $N$
coordinates. More sophisticated projection functions consider the
\emph{distribution} of the points, and project with the hyperplane which
preserves as much of the variance as possible (or, equivalently, reduces the
\emph{mutual information} between the points).

There are many techniques to find these hyperplanes, such as \emph{principle
  component analysis} (PCA) and \emph{autoencoding}. If the data are labelled,
these classes can also be taken into account during dimension
reduction~\cite{Oveisi.Oveisi.Erfanian.ea:2012}. However, since these
techniques are effectively ML algorithms in their own right, they suffer some of
the same constraints we're trying to avoid:

@@

\subsubsection{Sequences}

To handle input of \emph{variable} size, research attention has been given to
the handling of \emph{sequences}. This is a lossless approach, which splits the
input into fixed-size \emph{chunks} (e.g. splitting text into a sequence of
characters), which are fed into an appropriate ML algorithm one at a time. The
sequence is terminated by a sentinel; an ``end-of-sequence'' marker which, by
construction, is distinguishable from the data chunks. This technique allows us
to trade \emph{space} (the size of our input) for \emph{time} (the number of
chunks in a sequence).

@@

An alternative to introducing recurrent connections, is the use of
\emph{attention}. % TODO: cite
In a (typically feed-forward) ANN with attention, the whole input is provided to
the network, which produces some output as well as some ``attention''
information; the network is run again, but with the input adjusted using this
attention information (increasing and decreasing some of the input values); more
output is produced, along with new attention information, and so on. Such
networks do not need to ``remember'' long-distance relationships, whether by
recurrent connections or gated nodes; they instead use the same input over and
over again, but use their ``attention'' to emphasise those parts which should be
relevant to future iterations.

\subsubsection{Recursive Structure}

@@

@@
